---
title: "Charlotte Census data analysis"
author: Morgan Heinly
date: July 22, 2022
output:
      
    html_document:
        code_folding: none
        df_print: paged
        highlight: tango
        theme: flatly
        number_sections: yes
        toc: yes
        toc_depth: 2
editor_options:
  chunk_output_type: console


---

## load all necessary libraries


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE)


```

```{r}
library(tidycensus)
library(tidyverse)
library(remotes)
library(stringr)
library(scales)
library(plotly)
library(tigris)
options(tigris_use_cache = TRUE)
library(patchwork)
library(mapdeck)

```

## load in your api key for tidycensus

```{r}

api_key <-  '941ea2a2c4ffb8593729796459f5624ea053ea1c'

census_api_key(api_key,install = TRUE, overwrite = TRUE)

readRenviron("~/.Renviron")

```

## test a couple examples getting data with the API 

```{r}

options(tigris_use_cache = TRUE)

NC_med_house_income <- get_acs(
  state = "NC",
  geography = "tract",
  variables = "B19013_001",
  geometry = TRUE,
  year = 2020
)

head(NC_med_house_income) # view the data

```

## plot the results

```{r}

NC_med_house_income %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA,na.rm = TRUE) + 
  scale_fill_viridis_c(option = "viridis") +
  scale_fill_continuous(labels = scales::label_dollar(scale = .001,suffix = "K")) +
  theme_void() +
  ggtitle("2020 North Carolina median house hold income estimates by county") +
  theme(plot.title = element_text(hjust = 0.6,vjust = 1, size = 15)) + theme(legend.position = c(.35,.15))

```

## take a look at age and sex breakdown for 2019

```{r}


sex_breakdown <- get_estimates(
  geography = "state",
  state = "NC",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019
)

sex_breakdown <- sex_breakdown %>%
  filter(str_detect(AGEGROUP, "^Age"),
         SEX != "Both sexes") %>%
  mutate(value = ifelse(SEX == "Male", -value, value))

```

## visualize the data

```{r}


NC_pyramid <- ggplot(sex_breakdown, 
                       aes(x = value, 
                           y = AGEGROUP, 
                           fill = SEX)) + 
  geom_col(alpha = 0.9) + 
  theme_minimal() +
  scale_x_continuous(
    labels = ~ number_format(scale = .001, suffix = "k")(abs(.x)),
    limits = 400000 * c(-1,1)
  ) + 
  scale_y_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) + 
  scale_fill_manual(values = c("pink", "lightblue")) + 
  labs(x = "", 
       y = "2019 Census Bureau population estimate", 
       title = "Population structure in North Carolina", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")

NC_pyramid

```


## analyze move in / move out data in Charlotte from 2015 to 2020

## obtain access token for mapdeck

```{r}
token <- "pk.eyJ1IjoibW9yZ2FuaGVpbmx5IiwiYSI6ImNsNXNqMXdudjA3MjczZ21wb3RpMWs0MDQifQ.ZQFnbglE_9JBKOp5yrsxgQ"

```

## get the data

```{r}

CLT_flows <- get_flows(
  geography = "metropolitan statistical area",
  year = 2019,
  geometry = TRUE
)


CLT_flows <- CLT_flows %>%
  filter(str_detect(FULL1_NAME,"Charlotte-Concord"))

```

# top 25 cities moved to charlotte, nc

```{r}

top_moved_in <- CLT_flows %>%
  filter(!is.na(GEOID2), variable == "MOVEDIN") %>%
  slice_max(n = 25, order_by = estimate) %>%
  mutate(width = estimate / 500,
         tooltip = paste0(
           scales::comma(estimate * 5, 1),
           " people moved from ", str_remove(FULL2_NAME, "Metro Area"),
           " to ", str_remove(FULL1_NAME, "Metro Area"), " between 2015 and 2019"
         )
  )

```

## create moved in map from data

```{r Mapping inflows, echo=FALSE, message=FALSE, out.width = "100%",warning=FALSE}

####### create moved from map #################

G3 <- top_moved_in %>% 
  mapdeck(token = token, style = mapdeck_style("dark"), pitch = 45) %>% 
  add_arc(
    origin = "centroid1",
    destination = "centroid2", 
    palette = "plasma",
    stroke_width = "width",
    auto_highlight = TRUE,
    highlight_colour = "#0080FF80",
    tooltip = "tooltip"
  )

G3

```

## create moved out map from data

```{r Mapping outflows, echo=FALSE, message=FALSE, out.width = "100%",warning=FALSE}

top_moved_out <- CLT_flows %>%
  filter(!is.na(GEOID2), variable == "MOVEDOUT") %>%
  slice_max(n = 25, order_by = estimate) %>%
  mutate(width = estimate / 500,
         tooltip = paste0(
           scales::comma(estimate * 5, 1),
           " people moved to ", str_remove(FULL2_NAME, "Metro Area"),
           " from ", str_remove(FULL1_NAME, "Metro Area"), " between 2015 and 2019"
         )
  )

G4 <- top_moved_out %>% 
  mapdeck(token = token, style = mapdeck_style("dark"), pitch = 45) %>% 
  add_arc(
    origin = "centroid1",
    destination = "centroid2", 
    palette = "heat",
    stroke_width = "width",
    auto_highlight = TRUE,
    highlight_colour = "#0080FF80",
    tooltip = "tooltip"
  )

G4


```








